Developing Data Products - Week 2
=========================================================

####*Video 1.1-1.3 R Packages*

**A. R Packages Introduction**

- R packages are mechanisms for extending the basic functionality of R
- R packages are a collection of R functions and/or data objects
- R packages are organized in a systematic fashion to minimize inconsistency
- R packages are written by users/developers everywhere

**B. R Packages Locations**

- They live primarily on CRAN & Bioconductor (tools for biostatistics), but also on GitHub/Bitbucket
- Packages from CRAN/Bioconductor can be installed via "install.packages()" while GitHub ones can be installed via "install_github()" from the devtools package
- Packages do not have to be put in a central repository but it makes installation easier

**C. What's the Point?**

- There are reasons to not just "make some code available," and use packages instead:
    - Documentation is required
    - Centralized resources like CRAN
    - Knowledge that minimal standards like reliability/robustness met
    - Requires maintenance of the package to ensure integrity
    - Knowledge that the package will load properly
    
**D. Package Development Process**

1. Write some code in an R script file (.R)
2. Make the code available to others
3. Incorporate R script file into R package structure
4. Write documentation for user functions
5. Include supplemental materials (examples, demos, practice datasets, tutorials)
6. Package up and release!
7. Package will be submitted to CRAN or Bioconductor but also push source code to Github

- After submission, when problems are found with code, people will expect you to fix it and release a new version with the bugfix

**E. R Package Components**

- An R Package has:
    - a directory with the name of the R package
    - a DESCRIPTION file which has info about the package
    - R code (in the R/ sub-directory)
    - Documentation (in the man/ sub-directory)
    - NAMESPACE (optional, but do it)
    - Full requirements in writing R extensions

- The description file has:
    - Package: Name of package
    - Title: Full name of package
    - Description: longer description of package in one sentence (typically)
    - VersionL Version number (usually M.m-p format)
    - Author: name of original authors
    - Maintainer: Name + email of person who fixes problems
    - License: License for the source code
    - Depends: R packages that your package depends on
    - Suggests: Optional R packages that users may need but not strictly essential
    - Date: Release date of package
    - URL: package home page
    
- R Code:
    - Copy R code into the R/ sub-directory
    - There can be any number of files in this directory
    - Usually separate out files into logical groups
    - Code for all functions should be included here and here only
    
- The NAMESPACE File:
    - Used to indicate which functions are exported (public functions; the reason for package's existence)
    - Exported functions can be called by the user and are considered the public API
    - Non-exported functions cannot be called directly by the user (but the code can be viewed)
    - Hides implementation details from users and makes a cleaner package interface
    - You can also indicate what functions you import from other packages, allowing for your package to use other packages without making other packages visible to the user
    - Importing a function loads the package but does not attach it to the search list
    
- Documentation:
    - Documentation files (.Rd) placed in man/ sub-directory
    - Written in a specific markup language
    - Required for every exported function (another reason to limit exported function)
    - You can document other things like concepts, package overview, included data sets
    
**F. BUilding & Checking**

- R CMD build is a command-line program that creates a package archive file (.tar.gz)
- R CMD check runs a batter of tests on the package
- You can run "R CMD build" or "R CMD check" from the command-line using a terminal or command-shell application
- You can also run them from R using the system() function:

```{r}
# system("R CMD build newpackage")
# system("R CMD check newpackage")
```

- Checking runs a battery of tests including the existence of documentation, whether code can be loaded, runs documentation examples, and checks that documentation matches code
- Packages can only be put on CRAN if these battery of tests are passed

**G. Getting Started**

- The package.skeleton() function in the utils package creates a "skeleton" R package
- A "skeleton" R package means that a directory is created with the structure: (r/, man/), DESCRIPTION file, NAMESPACE file, documentation files
- If there are functions visible in the workspace, it writes R code files to the R/ directory and documentation stubs are created in man/
- This leaves the developer to fill in the rest!

**H. Package Development Summary**

1. Create a new directory with R/ and man/ sub-directories (just use package.skeleton())
2. Write a DESCRIPTION file
3. Copy R code into the R/ sub-directory
4. Write documentation files in man/ sub-directory
5. Write a NAMESPACE file with exports/imports
6. Build and check

####*Video 1.4-1.5 R Classes & Methods*

**A. Classes & Methods Overview**

- R classes & methods are implemented when there is a need to represent new data structures that do not exist yet in R (object oriented programming)
- Other OOP languages like C++, Java, & Python are not interactive unlike R (interactive means that you can script from the command line; the other languages are compiled and run)
- Chamber's vision was that beginners of R would first be "users" that utilize existing functions and, as their needs develop, eventually cross over and become "programmers" to develop additional functionality
- OOP is different in R than in most other languages

**B. Classes & Methods Styles**

- R classes and methods can be divided into 2 types:
    1) S3 classes/methods are included with version 3 of the S language, informal, and sometimes called old-style classes/methods
    2) S4 classes/methods are more formal/rigorous, included with R 1.4.0, and called new-style classes/methods
- For the foreseeable future, S3 & S4 classes/methods are separate systems and each system can be used fairly independently of the other
- Developers of new project are encouraged to use S4 (Bioconductor) but people still use S3 for "quick and dirty" development
- Code for implementing S4 classes/methods in R is in the "methods" package and can be loaded via library(methods)

**C. OOP in R**

- A class is a description of a thing and can be defined via setClass() in the methods package
- An object is an instance of a class and can be created via new()
- A method is a function that only operates on a certain class of objects
- A generic function is an R function which dispatches methods and encapsulates a "generic" concept such as plotting or predicting (the generic funtion itself does no computation)
- A method is the implementation of a generic function for an object of a particular class
- All objects in R have a class which can be determined by the class function

```{r}
class(1) # returns numeric
class(NA) # returns logical
class("foo") # returns character
```

- Classes go beyond the atomic classes (numeric, integer, character, logical, complex)

```{r}
x <- rnorm(100)
y <- x + rnorm(100)
fit <- lm(y ~ x)
class(fit) # an "lm" class
```

- S4 & S3 style generic functions look different but are conceptually the same
- Programmers can write new methods for an existing generic or create their own generics and associated methods
- If a data type does not exist in R, it can be defined with a new class along with the corresponding generics/methods

**D. S3 Classes**

- Simply calling a class name in R shows the namespace it comes from as well as the argument parameters passed in upon actual execution

```{r}
# show the environment and function arguments the following methods hail from
mean
print
```

- calling methods on a class shows its associated methods

```{r}
methods("mean") # what kind of methods are associated with the class "mean"?
```

- When a method is called on an object, R will attempt to execute a method specifically defined for that class. If none exists, the "generic" default function is called

```{r}
set.seed(2)
mean(rnorm(100)) # x is numeric but there is no mean method for "numeric" objects! call default mean function
```

- NEVER call methods directly, instead use the generic function and let the method be dispatched automatically (problem with S3; S4 does not allow direct calls)
- If new classes are created, you will probably have to write new methods for printing/showing, summary, & plotting

**E. S4 Classes**

- Why would you want to create a new class (expand beyond lists, numerics, data frames, etc)?
    - To represent new types of data that don't exist in R (gene expression)
    - New concepts/ideas that haven't been thought of yet (mixed-effects model, a sparse matrix)
    - To abstract/hide implementation details from the user

- A new class can be defined using the setClass function
    - At a minimum you need to specify the name of the class
    - You can also specify data elements that are called "slots"
    - You can then define methods for the class with the SetMethod function information about a class definition can be obtained with the showClass function
    
```{r}
# create a class called "polygon" that has an x & y numerical coordinate pair
library(methods)
setClass("polygon",
         representation(x = "numeric", y = "numeric"))

# create a plot method for the polygon class via setMethod()
setMethod("plot", "polygon",
          function(x, y, ...){
              plot(x@x, x@y, type = "n", ...)
              xp <- c(x@x, x@x[1])
              yp <- c(x@y, x@y[1])
              lines(xp, yp)
}) 
# note the slots of the polygon (the x & y coordinates) are accessed with the @ operator

# create an object from the polygon class
p <- new("polygon", x = c(1, 2, 3, 4), y = c(1, 2, 3, 1))
plot(p) # access the plot method which is NOT the default plot method, but rather is the one that was just created for the plot class
```

