<center> <h2>Getting & Cleaning Data - Week 2</h2> </center>
<center> <h3>---------------------------------------</h3> </center>

####*Video 1.1 Reading from mySQL*

**A. MySQL Overview**

- mySQL is free, open source, and widely used database software.
- Data are structured in a relational database that is structured as:
    - Databases
    - Tables within databases
    - Fields within tables
- Each row is called a record.
- mySQL can be run on the desktop or over a web server.

**B. Connecting to & Listing mySQL Databases**

- Connecting to a mySQL database requires installing the "RMySQL" package.
- A connection to a database is stored in a handler variable via the dbConnect() command with arguments:
    - MySQL(); indicates the kind of server being reached into.
    - user; an alias to the database
    - host; the URL to reach that database
- Queries into the databases are used via the dbGetQuery() command with arguments:
    - the database handle (standalone)
    - an SQL command in quotes (standalone)
- In order to save memory, the dbDisconnect() command should always be run, with the handle as the argument (will return TRUE to indicate that the connection is broken).

```{r echo=TRUE, eval=FALSE}
# store the query to the host website database in uscsDB
ucscDB <- dbConnect(MySQL(), user="genome", host="genome-mysql.cse.uscs.edu")

# execute the "show databases" command on the variable
result <- dbGetQuery(ucscDB, "show databases;")

# break the connection to ucscDB
dbDisconnect(ucscDB)
```
**C. Subsetting and Analyzing mySQL Databases**

- dbListTables() is a character vector storing all table names in the inputted database handle.

```{r echo=TRUE, eval=FALSE}
hg19 <- dbConnect(MySQL(), user="genome", db="hg19", host="genome-mysql.cse.uscs.edu")

#create a character vector storing all of the tables in the hg19 database
allTables <- dbListTables(hg19)
length(allTables)
allTables[1:5]
```

- dbListFields(<dbname>, <tablename>) lists all of the columns/fields in table <tablename> in <dbname>.
- dbGetQUery(<dbname>, <tablename>) tabulates records in the database in table <tablename> in <dbname>.

```{r echo=TRUE, eval=FALSE}
hg19 <- dbConnect(MySQL(), user="genome", db="hg19", host="genome-mysql.cse.uscs.edu")

# list every field/column in the database in a character vector
dbListFields(hg19, "affyU133Plus2") 

# count how many observations are in the database
dbGetQuery(hg19, "select count(*) from affyU133Plus2")
```

- dbReadTable(<dbname>, <tablename>) prints out the contents of <tablename> housed within <dbname>.

```{r echo=TRUE, eval=FALSE}
hg19 <- dbConnect(MySQL(), user="genome", db="hg19", host="genome-mysql.cse.uscs.edu")
# print out the table "affyU133Plus2" stored in the database handle hg19
affyData <- dbReadTable(hg19, "affyU133Plus2")
head(affyData)
```

- dbSendQuery() will go in the inputted database handle and execute SQL on it as a standalone argument.
- The SQL must simply be a valid, standard SQL query, and can include joins, fetches, etc.
- When using dbSendQuery(), a query is being sent into the database. fetch(queryname) must be executed and assigned to a variable. 
- Setting an n value equal to an integer in the fetch() command only returns the top n rows, useful in determining whether the query is returning the proper rows.
- After executing everything on a sent query, executing dbClearResult(<queryname>) removes the query from the server. 

```{r echo=TRUE, eval=FALSE}
hg19 <- dbConnect(MySQL(), user="genome", db="hg19", host="genome-mysql.cse.uscs.edu")

# in the hg19 database, go into the affyU133Plus2 table and return records where the misMatches values are between 1 and 3.
query <- dbSendQuery(hg19, "select * from affyU133Plus2 where misMatches between 1 and 3")

# actually fetch this inputted query and store it in a database called affyMis
affyMis <- fetch(query)

# return a quantile table, specifically subsetting the misMatches column from the dataframe affyMis
quantile(affyMis$misMatches)

# fetch the top 10 rows from the above query and clear the query
affyMisSmall <- fetch(query, n=10); dbClearResult(query)

# find the dimensions of the affyMisSmall fetch
dim(affyMisSmall) # returns 10 by 23, showing that there are 10 rows (by definition) and 23 fields
```

- Don't forget to close all sent queries and especially connections after they are done.

####*Video 1.2 Reading HDF5*

**A. HDF5 Introduction**

- HDF5 stores large data sets from a range of data types.
- The data format is hierarchical.
- Groups contain zero or more data sets/metadata.
    - A group header references a group name and attributes list
    - A group symbol table references a list of objects in a group
- Datasets are multidimensional array of data elements with metadata
    - Have a header with name, datatype, dataspace, and storage layout
    - Have a data array with the data
    
**B. Connecting to and Creating HDF5**

- HDF5 connection requires a bioconductor installation.
- The package is called biocLite("rhdf5")

```{r echo=TRUE, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite("rhdf5")
```

- The h5createFile() command creates an HDF5 file.

```{r echo=TRUE, eval=FALSE}
require(rhdf5)

# create an empty h5 file called example
created = h5createFile("example.h5")

# returns TRUE since this has been created
created
```

- The h5createGroup() command creates groups with the name of the inputted h5 file.
- hfls() lists the contents of the inputted h5 file.

```{r echo=TRUE, eval=FALSE}
# create a group called foo in the example h5 file
created = h5createGroup("example.h5", "foo")

# create a subgroup of foo called foobaa in the same file
created = h5createGroup("example.h5", "foo/foobaa")

# print out the group names and types of the file
h5ls("example.h5")
```

- The h5write() command writes the inputted data into the inputted group within the inputted file.
- inputting an index argument into h5write() will write the data into the inputted destination.

```{r echo=TRUE, eval=FALSE}
A = matrix(1:10, nr=5, nc=2)

# write the above matrix into the foo group of the example h5 file
h5write(A, "example.h5", "foo/A")

# write the numbers 12, 13, 14 into the top 3 rows of the first column of the A table in the foo group
h5write(12:14, "example.h5", "foo/A", index=list(1:3,1))
```

- Printing out the h5 file will show each data structure's group, name, type, class, and dimension.
- A new group can be created with the same name of the inputted data structure by simply executing the h5write() command.
- A data structure at the main level means that its group is just "/" (anything after indicates the subfolder and thus NOT main level).

```{r echo=TRUE, eval=FALSE}
# initialize a data frame
df = data.frame(1L:5L, seq(0, 1, length.out=5))

# write the df data frame to the example h5 file with the same name
h5write(df, "example.h5", "df")

# list out the groups
h5ls("example.h5")
```

- Groups from h5 can be read into a variable using the h5read() command.

```{r echo=TRUE, eval=FALSE}
# go into the foo/A portion of the example h5 file and store it in readA
readA = h5read("example.h5", "foo/A")
readA
```

####*Video 1.3-1.4 Reading from the Web (APIs)*

**A. Webscraping Introduction**

- Webscraping is the act of programmatically extracting data from the HTML code of websites.
- This may be against the terms of service of a website.
- Reading too much data in too quickly can get the IP address blocked.
- Connections to the internet need to be closed.

**B. Non-API Tools**

- Calling url() on a website will establish a connection.
- The HTML can be extracted using the readLines(conname) command on the connection conname.
- close(conname) is a command that closes the connection and can reduce slowdown problems.
- html can be parsed using the htmlTreeParse() command followed by xpathSapply() from the XML library.
- The "httr" package is another way to access html.
- Handles are a great way to access websites because they allow subsetting and other operations on the data.

```{r echo=TRUE, eval=FALSE}
# open a connection to Jeff Leek's Google citations
con = url("https://scholar.google.com/citations?user=HI-I6C0AAAAJ")

# extract the html code from this connection
htmlCode = readLines(con)

# close the connection
close(con)

htmlCode
```

**C. API Tools**

- An API (application programming interface) is a standard tool to access web data.
- Twitter is commonly used for this. A developer account is required and a unique application is allowed for each request that authenticates the developer user.

```{r echo=TRUE, eval=FALSE}
# Twitter Example
myapp = oauth_app("twitter", key="yourConsumerKeyHere", secret="yourConsumerSecretHere")
sig = sign_oauth1.0(myapp, token="yourTokenHere", token_secret="yourTokenSecretHere")

# allows a hit to access the JSON data stored on twitter
homeTL = GET("https://api.twitter.com/1.1/statuses/home_timeline.json", sig)
```

- IMPORTANT NOTE: the "foreign" package reads data from Octave, Stata, Minitab, etc.(e.g. read.spss)
