<center> <h2>Exploratory Data Analysis- Week 1</h2> </center>
<center> <h3>---------------------------------------</h3> </center>

####*Video 1.1 Principles of Analytic Graphics*

- There are general rules to using analytic graphics intended for use by others.

- Principle 1: Show Comparisons
    - Evidence for a hypothesis is always relative to another competing hypothesis
    - Always ask: "compared to what?"
    - This often involves comparing an experiment group's data values to a control group's
    
- Principle 2: Show causality, mechanism, explanation, systematic structure
    - What is the causal framework hypothesis for understanding the question?
    - Visualize the variables driving the final result in the way you visualize the causal relationship
    
- Principle 3: Show multivariate data (as much data on the plot as possible)
    - Multivariate data are more than 2 variables
    - The real world has multiple causal variables behind the dependent variable; show this!
    - Including all relevant variables reduces the possibility of omitted variable bias
    
- Principle 4: Integrate Evidence
    - Integrate words, numbers, images, diagrams
    - Data graphics should make use of many modes of data presentation in the final presentation
    - Don't let the tool drive the analysis, make the plot you want to make!
    
- Principle 5: Describe and document the evidence with appropriate labels, scales, sources, and so on
    - This gives objective context to the resulting graphics and results
    - Viewers can challenge the data by looking up the primary source making the process more scientific
    
- Principle 6: Content is King
    - The presentation depends on the quality, relevance, and integrity of the content
    - Aesthetic beauty is worthless without good data and logic tying it all together
    
####*Video 1.2-1.3 Exploratory Graphs*

**A. What is an Exploratory Graph?**

- Exploratory graphs are created by the user for his own use. 
- Graphing can be thought of serving 5 purposes in a data science pipeline:
    - To understand data properties
    - To find patterns in data
    - To suggest modelling strategies
    - To debug analyses
    - To communicate results
- Exploratory graphs are meant to aid in the first 4 of the above listed graphing purposes. They are used by the creator to understand the base hypothesis behind what is impacting the data.
- Exploratory plots are "quick and dirty" summarizations
- Characteristics of exploratory graphs:
    - They are made quickly
    - A large number are made
    - The goal is for personal understanding
    - Axes/legends are generally cleaned up, but not an immediate concern
    - Color/size are used primarily for underlying stastical information
    
**B. Methods of Exploratory Analysis - One Dimension**

- There are different methods of summarizing data that are useful for different situations and variable counts.
- One dimensional methods:
    - Key number summary (not a plot, states minimum, maximum, median, minimum, 1st & 3rd quartile)
    - Boxplots (graphically portrays the key number summary with a box around data center)
    - Histograms
    - Density plot
    - Barplot
    
```{R echo=TRUE, eval=FALSE}
# import air pollution data to answer question: do any counties exceed the standard air pollution?
if(!dir.exists(./data)) {dir.create(./data)}
# note this example can't be executed because RD Peng didn't include the website :<
pollution <- read.csv("./data/avgpm25.csv", 
                     colClasses=c("numeric", "Character", "factor", "numeric", "numeric"))
head(pollution)

# key number summary:
summary(pollution$pm25)

# box plot:
boxplot(pollution$pm25, col="blue")
abline(h=12) # draw in a line at a value of 12 to answer question of counties exceeding the standard air pollution
# histogram:
hist(pollution$pm25, col = "green")
hist(pollution$pm25, col = "green", breaks=100) # histogram but changes the number of bars
rug(pollution$pm25) # plots all of the actual points beneath the histogram

# barplot:
barplot(table(pollution$region), col="wheat", main="Number of Counties in Each Region") # plot the number of counties in each region by the categorical wheat column
```

**C. Methods of Exploratory Analysis - Multiple Dimensions**

- Two dimensions plotting methods
    - Multiple/overlayed 1-D plots (Lattice/ggplot2)
    - Scatterplots
    - Smooth scatterplots
    
- Three dimensions plotting methods
    - Overlayed/multiple 2-D plots; coplots
    - Use color, size, shape to add dimensions
    - Spinning plots
    - Actual 3-D plots (not that useful)

```{R echo=TRUE, eval=FALSE}
# import air pollution data to answer question: do any counties exceed the standard air pollution?
if(!dir.exists(./data)) {dir.create(./data)}
# note this example can't be executed because RD Peng didn't include the website :<
pollution <- read.csv("./data/avgpm25.csv", 
                     colClasses=c("numeric", "Character", "factor", "numeric", "numeric"))
head(pollution)
  
# boxplot:
boxplot(pm25 ~ region, data=pollution, col="red") # break down the air pollution by east/west region

# histograms:
par(mfrow=c(2, 1), mar=c(4, 4, 2, 1))
hist(subset(pollution, region=="east")$pm25, col="green") # plot the subsetted east
hist(subset(pollution, region=="west")$pm25, col="green") # plot the subsetted west

# scatterplot:
with(pollution, plot(latitude, pm25), col=region) # color differently by region to add dimension
abline(h=12, lwd=2, lty=2) # put in another abline at the critical value of 12

# multiple scatterplots:
par(mfrow=c(1, 2), mar=c(5, 4, 2, 1))
with(subset(pollution, region=="west"), plot(latitude, pm25, main="West")) # subset out west and plot
with(subset(pollution, region=="east"), plot(latitude, pm25, main="East")) # subset out east and plot
```

####*Video 1.4 Plotting Systems in R

**A. Base Plotting**

- This is the "artist palette" model in that it imitates the artist's painting process
- Start with a blank canvas and build it up
- Start with the plot function (or similar)
- Use annotation functions that will add/modify (test, lines, points, axis, etc)

- Pros and Cons of Base Plotting
    - Convenient, mirrors how most people would build a plot
    - Can't adjust a plot's fundamental nature once its started (margins can't be adjusted mid-way)
    - Difficult to translate to others what has happened (no graphical language) since plot is just a        series of ad hoc R commands.

```{r echo=TRUE, eval=TRUE}
# sample base plot
library(datasets)
data(cars) # pull in the mtcars dataset
with(cars, plot(speed, dist)) # from the mtcars dataframe, plot speed on x-axis, dist on y-axis
```

**B. The Lattice System**

- Called in the lattice library
- Plots are created with a single function call
- Most useful for conditioning types of plots (coplots): how y changes with x across levels of z
- Things like margins/spacing set automatically 
- Good for many plots on a screen
- Sometimes it is difficult to specify an entire plot in a single function call
- Annotation in plot is unintuitive
- Cannot "add" to plot once it is created

```{r echo=TRUE, eval=TRUE}
# sample lattice plot
library(lattice)
state <- data.frame(state.x77, region=state.region)
xyplot(Life.Exp ~ Income | region, data=state, layout=c(4, 1)) # from state DF, plot life exp as function of income across regions in 4 columns and 1 row
```

**C. The ggplot2 System**

- Splits the difference between base and lattice in that it doesn't need to be specified at once but still has many defaults. 
- Automatically deals with spacings, text, titles but allows adds to the plot
- More intuiutive than lattice
- Default mode makes many choices but they can be overwritten for user preference

```{r echo=TRUE, eval=TRUE}
library(ggplot2)
data(mpg)
qplot(displ, hwy, data=mpg) # plot variable "hwy" as a functino of "displ"
```

####*Video 1.5-1.7 Base Plotting Systems*

**A. The Packages Underlying Base Plotting**

- The core plotting and graphics engine in R is encapsulated in the following packages:
    - graphics: contains plotting functions for the "base" graphing systems, including plot, hist,           boxplot, etc
    - grDevices: contains all the code implementing the various graphics devices, including X11, PDF,        PostScript, PNG, etc

**B. The Process of Plotting**

- When making a plot, important considerations must be answered:
    - Where will the plot be made? ON the screen or in a file?
    - How will the plot be used?
        - Is the plot for viewing temporarily on the screen?
        - Will it be put on the web?
        - Will it be placed in a publication level paper?
        - Will it be in a presentation
    - Is there a large amount of data in the plot?
    - Will the graph need to be resized?
- What graphic system will allow for these goals: base, lattice, or ggplot2? These packages can NOT be mixed.

**C. Executing a Graph in Base R**

- Base graphics are used most commonly and utilize a very powerful system.
- There are two phases to creating a base plot:
    - Initializing a new plot
    - Annotating (adding to) an existing plot
- Calling plot(x, y) or hist(x) will launch a graphics device (if not open) and draw a new plot on the device.
- If the arguments to plot are not of some special class, the default method for plot is called.
- NOTE: the base graphics system has many parameters that can be tweaked and documented in "?par"

```{r echo=TRUE, eval=TRUE}
library(datasets)
hist(airquality$Ozone) # draw a new histogram using the defaults

with(airquality, plot(Wind, Ozone)) # plot a scatterplot of Ozone as a function of Wind

boxplot(Ozone ~ Month, airquality, xlab="Month", ylab="Ozone (ppb)") # box plot with axes labels
```

**D. Important Base Graphics Parameters/Functions**

- Base Plotting Functions:
    - plot: make a scatterplot, or other type of plot depending on the class of the plotting object
    - lines: add lines to a plot, given corresponding vectors of x and y values (connects dots) 
    - points: add points to a plot
    - text: add text labels to a plot using specified x, y coordinates
    - title: add annotations to x, y axis lables, title, subtitle, outer margin
    - mtext: add arbitrary text to the margins (inner or outer) of the plot
    - axis: adding axis ticks/labels

- Plot Specific Arguments:
    - pch: the plotting symbol (defaults to open circle)
    - lty: the line type (default is solid line), can be dashed, dotted, etc
    - lwd: the line width, specified as an integer multiple
    - col: the plotting color, speficied as a number, string, or hex code which are shown via colors()
    - xlab: character string for the x-axis label
    - ylab: character string for the y-axis label
    
- Environment Arguments (will impact ALL plots)
    - las: the orientation of the axis labels on the plot
    - bg: the background color
    - mar: the margin size
    - oma: the outer margin size
    - mfrow: number of plots per row, column (plots are filled row-wise)
    - mfcol: number of plots per row, column (plots are filled column-wise)
- Calling par("category") shows the default of the current styling category

```{r echo=TRUE, eval=TRUE}
library(datasets)

# graph 1 (base graph)
with(airquality, plot(Wind, Ozone))
title(main = "Ozone and Wind in New York City") # add a title

# graph 2 (color points)
with(airquality, plot(Wind, Ozone, main="Ozone and Wind in New York City"))
with(subset(airquality, Month == 5), points(Wind, Ozone, col="blue")) # color May points in blue

# graph 3 (add legend)
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City"), type="n") # create plot but do not place in the data
with(subset(airquality, Month == 5), points(Wind, Ozone, col="blue")) # color month 5 points blue
with(subset(airquality, Month == 5), points(Wind, Ozone, col="red")) # color non-month 5 points red
legend("topright", pch=1, col=c("blue", "red"), legend=c("May", "Other Months")) # create a legend indicating that May values are blue, all others are red

# graph 4 (add a regression line)
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City"), pch=20)
model <- lm(Ozone ~ Wind, airquality) # regress Ozone as a function of Wind
abline(model, lwd=2) # add a line to the plot, make width thicker

# graph 5 (multiple base plots)
par(mfrow=c(1, 2)) # overwrite the default and allow for two plots in 1 row, 2 columns
with(airquality, {
    plot(Wind, Ozone, main="Ozone and Wind") # graph #1
    plot(Solar.R, Ozone, main="Ozone and Solar Radiation") # graph #2
}
)

# graph 6 (multiple base plots with overarching title)
par(mfrow=c(1, 3), mar=c(4, 4, 2, 1), oma=c(0, 0, 2, 0))
with(airquality, {
    plot(Wind, Ozone, main="Ozone and Wild")
    plot(Solar.R, Ozone, main="Ozone and Solar Radiation")
    plot(Temp, Ozone, main="Ozone and Temperature")
    mtext("Ozone and Weather in New York City", outer = TRUE) # overarching title
})
```

**D. Base Plotting Demonstration**

```{r echo=TRUE, eval=TRUE}
x <- rnorm(100) # initialize a numeric vector
hist(x) # most basic histogram call with all defaults
y <- rnorm(100) # initialize a second numeric vector
plot(x, y) # plot y as a function of x
par(mar=c(2, 2, 2, 2)) # change the margin on each axis
plot(x, y)
plot(x, y, pch=20) # overwrite the fill
legend("topleft", legend="Data") # add a legend called "Data" to the upper left corner
abline(lm(y ~ x), lwd=3) # add the regression line of y as a function of x, thickening the line
z <- rpois(100, 2) # generate 100 random poissons with mean 2
par(mfrow = c(2, 1)) # plot in 2 rows, 1 column
plot(x, y, pch=20) # plot y as a function of x
plot(x, z, pch=19) # plot z as a function of x

g <- gl(2, 50) # generate 50 ones and twos
plot(x, y, type="n") # plot x and y but do not put in the data
points(x[g=="Male"], y[g=="male"], col="green") # initialize the x and y Male points to green
points(x[g=="Female"], y[g=="Female"], col="blue") # initialize the x and y Female points to blue
points(x[g=="female"], y[g=="female"], col="blue", pch=19) # plot these male and female points
```

####*Video 1.8 Graphic Devices in R*

**A. Graphic Device Introduction**

- A graphics device is something where you make a plot appear and includes:
    - A window on your computer
    - A PDF file
    - A PNG or JPEG file
    - A scalable vector graphics (SVG) file
- The most common place for a plot to be "sent" is the screen device. On Windows this is windows()

**B. How a Plot is Created**

- Method 1: Call a plotting function, the plot appears, annotate plot if necessary
- This method is useful for quick and dirty exploratory analysis

```{r echo=TRUE, eval=TRUE}
library(datasets)
with(faithful, plot(eruptions, waiting)) # make plot appear on screen device
title(main = "Old Faithful Geyser data") # annotate with a title
```

- Method 2: Explicitly launch a graphics device, call a plotting function, annotate plot, close graphics device with dev.off() (very important)!
- This method is used when you need to read the plot to disk to send to someone else

```{r echo=TRUE, eval=TRUE}
pdf(file=".myplot.pdf") ## open pdf device; create 'myplot.pdf' in working directory
with(faithful, plot(eruptions, waiting)) # create a scatter plot
title(main="Old Faithful Geyser data") # annotate plot
dev.off() # close the connection
```

**C. Graphics File Devices**

- Two basic types of file devices are vector and bitmap devices
- Vector formats (better for sparse plots):
    - pdf: useful for line-type graphs, resizes well, usually portable, better with sparse plots
    - svg: supports animation and interactivity
    - win.metafile
    - postcript: older format, resizes well, usually portable
- Bitmap formats (represents plots as pixels, better for dense data)
    - png: bitmapped format, good for line drawings with solid colors, good for plotting many points,        does not resize well
    - jpeg: good for photographs or natural scenes, uses lossy compression, not great for line drawings
    - tiff: supports lossless compression
    - bmp: a native Windows bitmapped format
    
**D. Multiple Open Graphics Devices**

- Open multiple graphics devices at once
- Plotting can only occur on one graphics device at a time
- The currently active graphics device can be found by calling dev.cur(), assigned integer >= 2
- dev.set() can overwrite the current device in use

- Copying Plots
    - dev.copy: copy a plot from one device to another
    - dev.copy2pdf: copy a plot specifically to a PDF file

```{r echo=TRUE, eval=TRUE}
library(datasets)
with(faithful, plot(eruptions, waiting)) # create plot on screen device
title(main="Old Faithful Geyser data") # add a main title
dev.copy(png, file="geyserplot.png") # copy plot to a png file in working directory
dev.off() # close the png device
```