<center> <h2>Exploratory Data Analysis- Week 2</h2> </center>
<center> <h3>---------------------------------------</h3> </center>

####*video 1.1-1.2 Lattice Plotting System*

**A. Lattice Plotting Overview**

- Useful for plotting...
    - many dimensions
    - highly dense plots
- The lattice package is built on the grid package
- The lattice plot, unlike base R plots which are built up like a canvas, is built in one function call and not easily altered
- Margins, spacing, and other formatting are handled automatically with typically sufficient defaults
- The lattice system is ideal for creating conditioning plots where the same plot is generated under many different conditions

- Fundamental lattice plots
    - xyplot: scatterplots
    - bwplot: box-and-whiskers plots ("boxplots")
    - histogram: histograms
    - stripplot: a boxplot with actual points
    - dotplot: plot dots on "violin strings"
    - splom: scatterplot matrix (like "pairs" in base plotting)
    - levelplot, contourplot: for plotting "image" data
    
- Contrasting Base R & the Lattice Package
    - AGAIN: you cannot mix functions from different plotting systems
    - Base graphics functions plot directly to the graphics device, while lattice graphics functions          return an object of class trellis (usually invisible to the user)
    - These trellis objects can, in principle, be stored in RAM variables 
    
**B. Scatterplots**

- The call for the scatterplot follows formula notation

```{r echo=TRUE, eval=FALSE}
# General Lattice scatterplot call:
xyplot(y ~ x | f * g, data)
# y is the y axis variable
# x is the x axis variable
# f & g are optional conditioning variables (categorical), breaks plot into panels as determined by layout() argument (specified to overwrite the default)
# data is the data frame or list from which the variables are referenced (if none is passed, the parent frame is used)

# Specific Lattice scatterplot calls:
library(lattice)
library(datasets)
# Simple call:
xyplot(Ozone ~ Wind, data=airquality) # go into the airquality dataframe in the global environment and plot Ozone as a function of Wind
# Same as above, but break down plot by month:
airquality <- transform(airquality, Month=factor(Month)) # convert Month into a categorical factor
xyplot(Ozone ~ Wind | Month, data=airquality, layout=c(5, 1))
```

**C. Lattice Panel Functions**

- Lattice functions have a panel function which controls what happens inside each panel of the plot
- The lattice package has default panel functions which can be overwritten to customize each panel
- Panel functions receive the x/y coordinates of the data points in their panel

```{r echo=TRUE, eval=TRUE}
# set up a simulated linear model:
set.seed(10)
x <- rnorm(100)
f <- rep(0:1, each=50)
y <- x + f - f * x + rnorm(100, sd=0.5) # simulate a linear model
f <- factor(f, labels=c("Group1", "Group2")) # convert 0 to group 1 and 1 to group 2 as a factor

# simple scatterplot:
xyplot(y ~ x | f, layout=c(2, 1))

# custom panel function with median y-axis line:
xyplot(y ~ x | f, panel=function(x, y, ...) { # "panel"" is an alias for each panel of the plot
    panel.xyplot(x, y, ...) # first call to plot 
    panel.abline(h=median(y), lty=2) # add a horizontal line at the median of the y axis
})

# custom panel function with median y-axis line:
xyplot(y ~ x | f, panel=function(x, y, ...) { # "panel"" is an alias for each panel of the plot
    panel.xyplot(x, y, ...) # first call to plot 
    panel.lmline(x, y, col=2) # overlay a simple linear regression line
})
```

####*Video 1.3-1.7 Plotting with ggplot2*

**A. What is ggplot2?**

- Implementation of the Grammar of Graphics by Leland Wilkinson (creates a language for visualization)
- Written by Hadley Wickham (as an Iowa State graduate student)
- "Shorten the distance from mind to page"

**B. The Workhorse : qplot()**

- Works much like the plot function in base R plotting (qplot -> "quick plot"").
- Looks for data in a data frame, similar to lattice, or in the parent environment
- Plots are made up of "aesthetics" (size, shape, and color) and "geoms" (points, lines)
- Factors are important for indicating categorical subsets of the data and they must be labeled
- The qplot() hides what goes on underneath, which is okay for most operations
- ggplot() is the core function and very flexible for doing things qplot() cannot

```{r echo=TRUE, eval=TRUE}
require(ggplot2)
str(mpg)

# ggplot2 "Hello, world!"
qplot(displ, hwy, data=mpg) # (x coord, y coord, data frame) 

# modifying aesthetics (classify by categorical variables)
qplot(displ, hwy, data=mpg, color=drv) # color the different kinds of drv categories differently
# NOTE: this different coloring of the drv variable occurred even though it is not a factor (it is a character vector)

# adding a geom
qplot(displ, hwy, data=mpg, geom=c("point", "smooth")) # plot the points & a fitted line with a confidence interval around it

qplot(hwy, data=mpg, fill=drv) # histogram of highway mileage, colored in based on the drv variable

qplot(displ, hwy, data=mpg, facets=.~drv) # produce the same scatterplot of hwy as a function of displ, but from left to right with each plot corresponding to a unique value of drv (this is like the color option, but instead of coloring the value, ggplot2 places them in different plots)

qplot(hwy, data=mpg, facets=drv~., binwidth=2) # produces vertically stacked histograms categorized based on the drv facet
```

**C. Delving Deeper & Customizing: ggplot()**

- Customizing should not be done in qplot(), turn to ggplot() instead.
- The components of a ggplot2 plot:
    - A data frame
    - aesthetic mappings: how data are mapped to color, size
    - geoms: geometric objects like points, lines, shapes
    - facets: used for conditional plots
    - stats: statistical transformations like binning, quantiles, smoothing
    - scales: what scale an aesthetic map uses (example: color males red, females blue)
    - coordinate system: communicating how numerical information maps to a plot
- The methodology of ggplot2 plotting:
    - The "artist's palette" model is what is implemented in ggplot2 (not qplot), in stark contrast to       the lattice system
    - Plots are built up in layers:
        - Plot the data
        - Overlay a summary
        - Metadata and annotation
- Call summary() on a ggplot stored variable to get basic information about the plot
- Base plot will print outliers if the coord_cartesian() option is included when data are truncated






