<center> <h2>Reproducible Research - Week 1</h2> </center>
<center> <h3>---------------------------------------</h3> </center>

####*Video 1.1 Coding Standards*

- Write code in a script
- Indent code 
- Limit the code's width (space down when it goes too far to the right)
- Limit the length of functions to execute just one fundamental activity (modularity)

####*Video 1.2 Markdown*

- Markdown is a very simplified markup language
- Markdown is a text-to-HTML tool that can easily convert one format to another
- Communicating different text formatting in Markdown...
    - Placing equal signs under text makes it look like a markdown file title
    - *this is italicized*
    - **this is bold**
    - ##this is a secondary heading
    - ###this is a tertiary heading
    - first item in list
    - second item in list
    1. first item in list
    2. second item in list
- Links can be inputted in MD by: [textHere](urlHere)
- Newlines require a double space after the end of a line

####*Video 1.3-1.4 R Markdown*

- Rmd integrates R code with a text readable document
- What is Markdown?
    - Allows one to focus on writing rather than formatting (formatting done easily)
    - Intuitive formatting elements
    - Easily converted to HTML and other formats
- What is R Markdown?
    - The integration of R code and markdown
    - Allows one to create R code meant for readers that contains live R code
    - R code is evaluated as part of the processing for markdown
    - Results from R code are inserted into markdown document
- How is R Markdown implemented?k
    - Can be converted to standard Markdown using the knitr R package
    - Easily implemented in R Studio
    - The resulting PDF or HTML should not be edited, just the underly RMD file should be
    
####*Video 1.5-1.8 Knitr*

**A. Literate Statistical Programming**

- Literate statistical programming promotes the integration of machine code and human readable text in one file (e.g. R Markdown)
- An empirical project is a stream of text & code (therefore, writeups alter between text and code)
- Literate programs can be "weaved" into human-readable documents (e.g. a knitted RMD file) and "tangled" into machine-readable documents (code files)

- Literate Programming Pros:
    - text and code are in one place and in a logical order
    - data and results automatically update with new code
    - program breaks when an error is introduced, so bugs do not linger
    
- Literate Programming Cons:
    - documents with a lot of code are hard to read
    - processing of documents can be very slow

**B. How Can a Project be Made Reproducible?**

- Commit to do it from the start
- Use version control (Git)
- Use software that can be coded
- Don't save the output, save the raw data and the programs that convert those files to output

**C. What is Knitr?**

- Knitr is a CRAN stored R package
- Knitr supports RMarkdown, LaTex, and HTML as documentation languages and can export them to PDF or HTML
- Knitr is built into RStudio
- Using knitr requires knowledge of either RMarkdown, LaTex, or HTML (JHU uses RMarkdown)

- Pros of knitr:
    - Knitr is good for manuals (how to use a piece of software)
    - Short to medium length technical documents
    - Tutorials
    - Reports
    - Data preprocessing documents/summaries
    
- Cons of knitr:
    - Bad for very long research articles and time consuming computations (slow computation       time)
    - Documents requiring precise formatting (aesthetics limited to what RMarkdown can do)
    
**D. The Mechanics of Knitr**

- Code chunks begin with "'''{r}" and end with "'''"
- Code chunks can have names, just passing them after the "r" declaration at start

```{r CODECHUNK1}
# this R code block is called "CODECHUNK1"
print("CODECHUNK1 Executes")
```
   
- Setting "results=asis" prints the output verbatim instead of rendering it into HTML 
- Setting "echo=FALSE" after R declaration indicates that the code will not be printed out
- Setting "eval=FALSE" prints out the code but does not print the output

```{r, echo=FALSE}
# this R code block is called "CODECHUNK1"
print("This will not be printed")
```

- opts_chunks$set is a function that can set global arguments for the chunks of code in an RMarkdown file (must be defined in the initial code chunk)
- fig.height and fig.width accept numerical values to communicate the size of outputted figures
- Caching can occur on a chunk-by-chunk basis by declaring "chunk=TRUE" which evaluates the code the first time, but will load the code off of disk if it is run again and that chunk is unaltered
- Note that a code chunk may not be cachable if it is directly related to other code in uncachable ways
    