<center> <h2>R Programming - Week 3</h2> </center>
<center> <h3>---------------------------------------</h3> </center>

####*Video 1.1 The str Function*

- The str function is the most important function in all of R.
- The str function compactly displays the summary of R objects.
- str provides roughly one lines of output per basic object.

```{r echo=TRUE, eval=TRUE}
str(str) # shows that the input to the str function is an object

str(lm) # shows the arguments to the lm function

x <- rnorm(100, 2, 4)
summary(x) # shows the minima and maxima and other summary statistics
str(x) # shows the range of the command in generating x, as well as the first batch of random variates

f <- gl(40, 10) # generate 40 levels 10 times each
str(f) # says this is a factor object, how many loevels, and how many times each is repeated

library(datasets)
str(airquality) # says that this object is a data frame, how many observations/characteristics, and some of the first observations

m <- matrix(rnorm(100), 10, 10)
str(m) # shows the dimensions of m, the class of m, and the first part of the first column of m
```

####*Video 1.2 - 1.4 Simulation

**A. Generating Random Numbers**

- Functions for probability distributions in R:
    - rnorm: generate random Normal variates with a given mean and sd
    - dnorm: evaluate the Normal probability density at a (vector of) points
    - pnorm: evaluate the cumulative distribution function for a normal distribution
    
- Probability distribution functions typically have four functions associated with them:
    - d for density
    - r for random number generation
    - p for cumulative distribution
    - q for quantile function

```{r echo=TRUE, eval=TRUE}
dnorm(x, mean = 0, sd = 1, log = FALSE)
pnorm(q = 0.5, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
qnorm(p = 0, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
rnorm(n = 50, mean = 0, sd = 1)

# standard random normal examples
dnorm(x = 0) # the height of the standard normal at x = 0 is 0.3989
qnorm(p = 0.5) # the location of x where half of the density to the left is 0
pnorm(q = 0) # when x is equal to zero, half of the density is to the left

x <- rnorm(10) # generate 10 standard normals randomly
x <- rnorm(10, 20, 2) # generate 10 random normals with mean 20 and sd of 2
summary(x)
```

- Setting the random number seed with set.seed ensure reproducibility by generating the same random normals at each run.
- set.seed(int) sets a seed identified by the integer passed in (arbitrarily chosen).

```{r echo=TRUE, eval=TRUE}
set.seed(1)
rnorm(5)
rnorm(5) # generates a different batch, you didn't call set.seed() again
set.seed(1)
rnorm(5) # same randoms generate as initial when the seed was set

rpois(10, 1) # 10 Poissons with a mean of 1
rpois(20, 30) # 20 Poissons with a mean of 30
ppois(2, 2) # probability a Poisson with mean of 2 is less than or equal to 2
ppois(4, 2) # probability a Poisson with mean of 2 is less than or equal to 4
ppois(6, 2) # probabilit a Poisson with mean of 2 is less than or equal to 6
 ```

**B. Simulating a Linear Model**

```{r echo=TRUE, eval=TRUE}
# simulate y as a function of x where a y-intercept of .5 and a slope of 2 are assumed
set.seed(20)
x <- rnorm(200) # generate 200 standard random normals
e <- rnorm(100, 0, 2) # generate 100 random normals with mean 0 and sd of 2 (noise)
y <- 0.5 + 2 * x + e # y is a function of 2 times x, plus 0.5, plus white noise
plot(x, y) 
```

**C. Random Sampling**

- The sample function allows random sampling of elements in an inputted vector.
- If the number of trials is left blank, a permutation is given.
- Unless overwritten, the default is to sample without replacement.

```{r echo=TRUE, eval=TRUE}
sample(1:10, 4) # sample 4 elements from 1 through 10 without replacement
sample(letters, 5) # sample 5 letters without replacement
sample(1:10) # sample every number 1 through 10 in random order (permutation)
sample(1:10, replace = T)  # sample 10 elements but with replacement (repeats possible)
```

####*Video 1.5 The R Profiler#

**A. Profiling Overview**

- The R Profiler is a tool to assess programs that are taking a long time.
- Profiling systematically examines how much time is spent running a program.
- In order to speed up code, you need to know where time is being spent.
- Profiling is useful for testing code that runs 1000s of times in loop: can it be made more efficient???
- Design first, then optimize.

**B. system.time()**

- system.time() takes an arbitrary R expression as input and returns evaluation time.
- Returns an object of class proc_time
- Time of execution can be divided into two parts:
    - User time (time charged to the CPU)
    - Elapsed time (time spent waiting around before CPU execution)
- Elapsed time is greater than user time if CPU is just waiting around.
- User time is greater than elapsed time if machine has multiple cores.
- R defaults to using 1 core, but can be overwritten using parallel computing packages.
- system.time() is only useful if you know where the problem is.

**C. The R Profiler**

- The Rprof() starts the profiler in R.
- The summaryRprof() summarizes the results of Rprof().
- Rprof() keeps track of the function call stack and says how long different function executions take.
- Subtracting lower level function run time from upper level, in nested functions, is necessary to tell where real time's spent(time can be spent in multiple functions at once).
- If R code calls C/Fortran, this code is not profiled.

